Dish: Distribution-oblivious Shell Scripting

### 2019-11-22 pldi/draft.tex

\newcommand{\eg}{{\em e.g.}, }
\newcommand{\ie}{{\em i.e.}, }
\newcommand{\etc}{{\em etc.}\xspace}
\newcommand{\vs}{{\em vs.} }
\newcommand{\cmpn}{compartmentalization} 
\newcommand{\heading}[1]{\vspace{4pt}\noindent\textbf{#1}\enspace}
\newcommand{\ttt}[1]{\texttt{\small #1}}
\newcommand{\ttiny}[1]{\texttt{\scriptsize #1}}
\newcommand{\spol}[1]{\scriptsize{\sc#1}}
\newcommand{\pol}[1]{\texttt{\small {\color{purple}#1}}}
\newcommand{\rf}[1]{\ref{#1}}
\newcommand{\wka}{\ttt{a\textsubscript{1}}}
\newcommand{\wkq}{\ttt{q\textsubscript{1-4}}}

@@

% A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
\unix provides several ways for succinct program composition.
Central among them is the \emph{pipe}, a primitive that passes the output of one program as input to the next.
These programs produce output and consume input concurrently, and possibly at different rates, with the \unix kernel facilitating program scheduling and communication synchronization behind the scenes.
Intermediate programs, termed commands, can also be developed in different programming languages, as long as they conform to a common interface.
\begin{wrapfigure}{r}{0.06\textwidth}
  \vspace{-12pt}
  % \begin{center}
    % \includegraphics[width=0.48\textwidth]{gull}
    \includegraphics[width=0.06\textwidth]{\detokenize{./figs/dish_ex1.pdf}}
  % \end{center}
  \vspace{-30pt}
\end{wrapfigure}
This interface exposes a stream of contiguous character lines.
A few characters are special---for example, the newline (\textsc{NL}) character delineates an element of the stream, and the end-of-file (\textsc{EOF}) character signals the end of a stream.
As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
% the input stream (stdin), the output stream (stdout), and the generating errors or diagnostics to the standard error stream (stderr)
% generating errors or diagnostics to the standard 
% Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
% TODO: \kk{We have to mention character vs line based parallelism somewhere, and that we choose to see lines as the data quantum}

@@

\section{Distributability Classes}

\subsection{Comnands}
\label{bg:cmd}

% Sets vs. Classes

\begin{table}[h]
\center
\footnotesize
\setlength\tabcolsep{3pt}
\caption{
  \footnotesize{
    \textbf{Distributability Classes}.
    Broadly, \unix commands can be broken down into to five classses.
  }
}
\begin{tabular}{l @{\extracolsep{\fill}} lll}
\toprule
Class                           & Example Commands        & POSIX            & Core-utils\\
\midrule
Stateless                       &                         &                  &           \\
Pure                            &                         &                  &           \\
(Distr.) File-System            &                         &                  &           \\
Side-effectful                  &                         &                  &           \\
Irreversibly Side-effectful     &                         &                  &           \\
% \midrule
% Shell PL Constructs             &                         &                  \\
% \etc                            &                         &                  \\
\bottomrule
\end{tabular}
\label{tab:classes}
\end{table}

Define distributability 
Explain distributability classes, based on how

\heading{Preliminaries}

\heading{POSIX, GNU Core-utils, and beyond}

\heading{Stateless}

\heading{Pure}

\heading{File-system}

\heading{Side-effectful}

\heading{Irreversably Side-effectful}

\subsection{The Shell Language}
\label{bg:shell}

\heading{Implicits}
Expansion
Implicits
Context

\heading{Language Constructs}

Even without these, the Unix pipeline language is powerful enough to capture complex pipelines.

We can encode powerful

\section{Dish Overview}

>>>>

\newcommand{\eg}{{\em e.g.}, }
\newcommand{\ie}{{\em i.e.}, }
\newcommand{\etc}{{\em etc.}\xspace}
\newcommand{\vs}{{\em vs.} }
\newcommand{\cmpn}{compartmentalization} 
\newcommand{\heading}[1]{\vspace{4pt}\noindent\textbf{#1}\enspace}
\newcommand{\ttt}[1]{\texttt{\small #1}}
\newcommand{\ttiny}[1]{\texttt{\scriptsize #1}}
\newcommand{\tti}[1]{\texttt{\scriptsize #1}}
\newcommand{\spol}[1]{\scriptsize{\sc#1}}
\newcommand{\pol}[1]{\texttt{\small {\color{purple}#1}}}
\newcommand{\rf}[1]{\ref{#1}}
\newcommand{\wka}{\ttt{a\textsubscript{1}}}
\newcommand{\wkq}{\ttt{q\textsubscript{1-4}}}

\newcommand{\cn}[1]{\mbox{\textcircled{\footnotesize #1}}}
\newcommand{\tcn}[1]{\mbox{\textcircled{\scriptsize #1}}}

\newcommand{\pur}{\cn{\textsc{P}}\xspace}
\newcommand{\sta}{\cn{\textsc{S}}\xspace}
\newcommand{\dfs}{\cn{\textsc{F}}\xspace}
\newcommand{\sid}{\cn{\textsc{E}}\xspace}
\newcommand{\irr}{\cn{\textsc{I}}\xspace}

\newcommand{\tpur}{\tcn{\textsc{P}}\xspace}
\newcommand{\tsta}{\tcn{\textsc{S}}\xspace}
\newcommand{\tdfs}{\tcn{\textsc{F}}\xspace}
\newcommand{\tsid}{\tcn{\textsc{E}}\xspace}
\newcommand{\tirr}{\tcn{\textsc{I}}\xspace}

@@

% A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
\unix provides several ways for succinct program composition.
Central among them is the \emph{pipe}, a primitive that passes the output of one program as input to the next.
These programs produce output and consume input concurrently, and possibly at different rates, with the \unix kernel facilitating program scheduling and communication synchronization behind the scenes.
Intermediate programs, termed commands, can also be developed in different programming languages, as long as they conform to a common interface.
\begin{wrapfigure}{r}{0.06\textwidth}
  \vspace{-12pt}
  % \begin{center}
    % \includegraphics[width=0.48\textwidth]{gull}
    \includegraphics[width=0.06\textwidth]{\detokenize{./figs/dish_ex1.pdf}}
  % \end{center}
  \vspace{-30pt}
\end{wrapfigure}
This interface exposes a stream of contiguous character lines.
A few characters are special---for example, the newline (\textsc{NL}) character delineates an element of the stream, and the end-of-file (\textsc{EOF}) character signals the end of a stream.
As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
% the input stream (stdin), the output stream (stdout), and the generating errors or diagnostics to the standard error stream (stderr)
% generating errors or diagnostics to the standard 
% Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.

@@

\section{Distributability Classes}
\label{distributability}

To have any hope of distributing \unix pipelines, \sys first needs to understand the distribution characteristics of their intermediate commands.
This understanding is encoded in a command's \emph{distributability} class.
Informally, a distributability class represents the level of synchronization required by copies of a command executing in parallel.
It focuses on characteristics that are important for distribution, rather than all the observable behavior of a command.
\sys leans towards a few coarse classes rather than many detailed ones---among other reasons, to simplify their use by command developers.
% Make the connections that this is like a type, that doesn't capture everything, but is good-enough

This section starts by defining these classes, paired with a distributability study of the commands found in standard libraries such as {\sc POSIX} and GNU Coreutils~\sx{cmd}.
Building on this study, it presents a DSL that enables command classification by its developers~\sx{ext}.
In return, \sys uses this DSL to tag commands in the standard libraries and generate their wrappers presented in later sections.

\subsection{Distributability of Standard Libraries}
\label{cmd}

% nv TODO: this is a hierarchy
Broadly, shell commands can be split into five major classes (summarized in Tab.~\ref{tab:classes}) with respect to their distribution characteristics, depending on how they interact with state.
These classes are ordered in ascending difficulty of distribution:
  later classes require more effort.
In this order, each class can be thought as a subset of the next one---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without seeing any performance improvements).

\begin{table}[t]
\center
\footnotesize
\setlength\tabcolsep{3pt}
\caption{
  \footnotesize{
    \textbf{Distributability Classes}.
    Broadly, \unix commands can be broken down into to five classes.
  }
}
\begin{tabular}{l @{\extracolsep{\fill}} llll}
\toprule
Class                           &  Key    & Example Commands                            & Coreutils       & POSIX       \\
\midrule
Stateless                       & ~\tsta  & \tti{tr},   \tti{cat},    \tti{grep}        &  22 (21.1\%)     &           \\  % 22
Pure                            & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  21 (20.1\%)     &           \\  % 21
(Distr.) File-System~           & ~\tdfs  & \tti{cp},   \tti{ln},     \tti{rm}          &  27 (25.9\%)     &           \\  % 27
Side-effectful                  & ~\tsid  & \tti{env},  \tti{chroot}, \tti{whoami}      &  25 (24\%)       &           \\  % 25
Irreversible                    & ~\tirr  & \tti{exit}, \tti{lpr},    \tti{reboot}      &  5  (0.4\%)      &           \\  % 5
% \midrule
% Shell PL Constructs             &                         &                  \\
% \etc                            &                         &                  \\
\bottomrule
\end{tabular}
\label{tab:classes}
\end{table}

% \heading{Preliminaries}
% \heading{POSIX, GNU Core-utils, and beyond}

\heading{Stateless Commands}
The first class contains stateless commands~\sta that operate on individual elements of the stream, without maintaining state across invocations.
These are commands that can be expressed as a purely functional map---\eg \ttt{basename} removes a path prefix from a string and \ttt{grep} filters out individual lines.
At times, stateless commands may operate as functions that produce multiple elements---\eg when they insert a {\sc NL} token.
Workloads that use only stateless commands are trivial to parallelize:
  they do not require any synchronization to maintain correctness, nor caution about where to split.
 % map :: (a -> b) -> [a] -> [b]

Many of these commands (about 1/3 of coreutils' \sta) are stateless \emph{within} an element (\eg \ttt{tr} transliterates characters, one at a time).
This property could allow further parallelization within a single line element, a feature that might seem of limited use because these commands are not particularly expensive computationally precisely due to their narrow focus.
However, it is useful for cases with very large stream elements (long lines) such the \ttt{.fastq} format used in bioinformatics pipelines.

\heading{Pure Commands}
The second class contains pure commands \pur.
These commands respect functional purity, returning the same results for the same inputs, but maintain internal state across their entire pass.
The details of this state and its propagation during piecewise element processing affects their distributability characteristics.

Some commands are easy to parallelize, because they maintain trivial state and are commutative---\eg \ttt{wc}, which simply maintains a counter.
Others, such as \ttt{sha1sum}, maintain more complex state that is not applied commutatively.

Often these commands do not operate in an online fashion, but need to block until the end of a stream---a typical example of this is \ttt{sort}, which cannot start before the last line of input has been consumed.
Non-streaming constraints affect task parallelism, but not data parallelism:
  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques.

\heading{File-system Commands}
The third class contains commands that access the file-system~\dfs.
This is the largest class, as the file (and file-system) abstraction is central to the \unix design and philosophy~\cite{unix}.
In fact, \unix uses the file-system as a proxy to several file-unrelated operations---\eg access control and device driving is grafted upon the file-system.

% TODO: mention that some commands fail to comply with the interface---for example, csplit could be split followed by a redirection to files.
Commands in \dfs are distributable for many reasons, as evidenced by a long history of distributed \unix file-system clones~\cite{catalogue}.
Some of these offer the semantics of the \unix file-system---hierarchies, protection bits---and a POSIX virtualization layer that enables support for the commands in \dfs.
The vast majority of these commands are control-plane operations that act on identifiers~\sx{bg}---for example, \ttt{touch} for creating a file or \ttt{chgrp} for changing group ownership.
% these operate by changing an inode
Data-plane operations on files go explicitly through stream-redirection operations~\sx{bg}.
Operations using different identifiers pose no problems;
  operations sharing a single identifier might have problems---but these risks exist even in the non-distributed case.
While such pathological cases are possible, they are rare enough---and have been rare enough since \unix's inception~\cite[\S3.6]{unix}).

% \unix itself does not offer any guarantees when a file is accessed concurrently on a single computer.

\heading{Side-effectful Commands}
Another large class of commands are ones that have side-effects to memory~\sid.
Because of timesharing and multi-tasking---\ie the fact that many users can operate multiple terminals on single \unix machine---a subset of these commands are devoted to accessing and sharing memory such as environment variables and shell parsing flags.

A command in \sid will require some form of concurrency control to distribute property.
However, because its side-effects stay within the distributed \unix environment, they can be manipulated using---including optimistic and multi-versioning schemes that can roll-back side-effects.

The vast majority of these commands (a ratio of 21:4) only \emph{read} values, and usually values that are never affected by the user.
For example, \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands that interface with kernel- or hardware-generated information that is not writable by user-space scripts.
Even if \sid commands are protected with transactions
  these benefits are even greater for read-only or read-mostly accesses.

\heading{Irreversibly Side-effectful Commands}
Finally, some commands have side-effects outside \sys---for example, by printing a file with \ttt{lpr} or restarting with \ttt{reboot}.
Such commands are not distributable---they have to refer to a single location---and have to be guarded with pessimistic transaction control mechanism.
